---
title: "exam_pmm"
author: "Schloessmann"
date: "2026-02-05"
output: html_document
---

# 0.0 load libraries

```{r}
library(deSolve)
library(lattice)
library(growthrates)
library(ggplot2)
```

# 1.0 Load data
```{r}
picea_growth <- read.delim("picea.txt", stringsAsFactors=TRUE)
```

## 1.1 assign datatypes
```{r}
picea_growth$location <- as.factor(picea_growth$location)
picea_growth$tree <- as.factor(picea_growth$tree)
picea_growth$volume <- as.double(picea_growth$volume)
```

# 2.0 Subsetting the areas
```{r}
area1 <- subset(picea_growth, location == 2) # altitude above 1000m, cold temperatures
area2 <- subset(picea_growth, location == 7) # altitude below 1000m, temperate climate like in Harz region
```

```{r}
head(picea_growth)
```

# 3.0 plot height and dbh by location
```{r}
plot(
  picea_growth$height,
  picea_growth$dbh,
  xlab = "height",
  ylab = "dbh",
  col = picea_growth$location
)
```

```{r}
library(ggplot2)

ggplot(picea_growth, aes(x = age.base, y = height)) +
  # Punkte für Area 1
  geom_point(
    data = subset(picea_growth, location == 2),
    aes(color = "area 1"),
    shape = 1
  ) +
  # Punkte für Area 2
  geom_point(
    data = subset(picea_growth, location == 7),
    aes(color = "area 2"),
    shape = 1
  ) +
  # Farben manuell festlegen
  scale_color_manual(
    name = "Legend",
    values = c(
      "area 1" = "magenta",
      "area 2" = "black"
    )
  ) +
  labs(
    x = "Age",
    y = "Height [m]",
    #title = "Picea abies height growth"
  ) +
  theme_minimal()
```

# 4.0 Creating model for area1 (black/the higher altitude)
```{r}
max(area1$height)
```


```{r}
x <- area1$age.base
y1.0 <- grow_logistic(x, c(y0=0.1, mumax=0.05, K=38.1))[,"y"]
y1.1 <- grow_gompertz(x, c(y0=0.1, mumax=0.05, K=38.1))[,"y"]
```



```{r}
ggplot(area1, aes(x=age.base, y=height)) +
  geom_point(shape=1, col="magenta") +
  geom_line(aes(x=x, y=y1.0), col="red") +
  geom_line(aes(x=x, y=y1.1), col="green")
```
```{r}


ggplot(area1, aes(x = age.base)) +
  geom_point(aes(y = height, color = "area 1"), shape = 1) +
  geom_line(aes(y = y1.0, color = "Logistic model")) +
  geom_line(aes(y = y1.1, color = "Gompertz model")) +
  scale_color_manual(
    name = "Legend",
    values = c(
      "area 1" = "magenta",
      "Logistic model" = "red",
      "Gompertz model" = "green"
    )
  ) +
  labs(
    title = "First try for growthmodels for area 1",
    x = "Age",
    y = "Height [m]"
  ) +
  theme_minimal()
```

 
# 5.0 Model calibration
```{r}
p   <- c(y0 = 0.1, mumax = 0.05, K = 38.1)
fit1.0 <- fit_growthmodel(grow_logistic, p=p, time=area1$age.base, 
                       y=area1$height)

fit1.1 <-fit_growthmodel(grow_gompertz, p=p, time=area1$age.base, 
                       y=area1$height)
?fit_growthmodel
```

```{r}
fit1.0@fit$par
```

```{r}
fit1.1@fit$par
```


```{r}
res1.0  <-as.data.frame(predict(fit1.0))
res1.1 <- as.data.frame(predict(fit1.1))

```

```{r}
ggplot(area1, aes(x = age.base)) +
  geom_point(aes(y = height, color = "trees area 1"), shape = 1) +
  #geom_line(data = res1.0, aes(x = x, y = y, color = "logistic")) +
  geom_line(data = res1.1, aes(x = x, y = y, color = "growth model")) + #Gompertz
  scale_color_manual(
    name = "Legend",
    values = c(
      "trees area 1" = "magenta",
    #  "logistic" = "red",
      "growth model" = "green"
    )
  ) +
  labs(
    x = "Age",
    y = "Height [m]"
  ) +
  theme_minimal()
```

```{r}
head(fit1.0@fit$residuals)
```
## 5.1 Check the sum of squared residuals
```{r}
sum(fit1.0@fit$residuals^2)
```

```{r}
sum(fit1.1@fit$residuals^2)
```
In case 1.0 the sum of the squared residuals is higher, that means, that the residuals are farer away from the fitted curve in case 1.0 compared to 1.1(gompertz)

# _____________________________________________________

# 6.0 parameters for area2

```{r}
max(area2$height)
```

```{r}
x2 <- area2$age.base
y2.0 <- grow_logistic(x2, c(y0=0.1, mumax=0.05, K=21.6))[,"y"]
y2.1 <- grow_gompertz(x2, c(y0=0.1, mumax=0.05, K=21.6))[,"y"]
```

## 6.1 Plotting the parameters for area2 with logistic and Gompertz 
```{r}
ggplot(area2, aes(x = age.base)) +
  geom_point(aes(y = height, color = "area 2"), shape = 1) +
  geom_line(aes(y = y2.0, color = "Logistic model")) +
  geom_line(aes(y = y2.1, color = "Gompertz model")) +
  scale_color_manual(
    name = "Legend",
    values = c(
      "area 2" = "black",
      "Logistic model" = "red",
      "Gompertz model" = "green"
    )
  ) +
  labs(
    title = "First try for growthmodels for area 2",
    x = "Age",
    y = "Height [m]"
  ) +
  theme_minimal()
```

# 7.0 fitting 
```{r}
p2  <- c(y0 = 0.1, mumax = 0.05, K = 26.1)
fit2.0 <- fit_growthmodel(grow_logistic, p=p2, time=area2$age.base, 
                       y=area2$height)

fit2.1 <-fit_growthmodel(grow_gompertz, p=p2, time=area2$age.base, 
                       y=area2$height)
```

```{r}
fit2.0@fit$par
```

```{r}
fit2.1@fit$par
```

# 7.1 Creating dataframes with fitted values

```{r}
res2.0  <-as.data.frame(predict(fit2.0))
res2.1 <- as.data.frame(predict(fit2.1))
```

```{r}
ggplot(area2, aes(x = age.base)) +
  geom_point(aes(y = height, color = "trees area 2"), shape = 1) +
 # geom_line(data = res2.0, aes(x = x2, y = y, color = "logistic")) +
  geom_line(data = res2.1, aes(x = x2, y = y, color = "growth model")) +
  scale_color_manual(
    name = "Legend",
    values = c(
    "trees area 2" = "black",
     # "logistic" = "red",
      "growth model" = "green"
    )
  ) +
  labs(
    x = "Age",
    y = "Height [m]"
  ) +
  theme_minimal()
```

## Check residuals

```{r}
sum(fit2.0@fit$residuals^2)
```

```{r}
sum(fit2.1@fit$residuals^2)
```

# 8.0 Comparing

```{r}
ggplot(area2, aes(x = age.base)) +
 # geom_point(aes(y = height, color = "trees area 2"), shape = 1) +
  geom_line(data = res1.1, aes(x = x, y = y, color = "growth model a1")) +
  geom_line(data = res2.1, aes(x = x2, y = y, color = "growth model a2")) +
  scale_color_manual(
    name = "Legend",
    values = c(
    "growth model a1" = "magenta",
     # "logistic" = "red",
      "growth model a2" = "black"
    )
  ) +
  labs(
    x = "Age",
    y = "Height [m]"
  ) +
  theme_minimal()
```

```{r}
ggplot() +
  # Points for Area 1
  geom_point(data = area1, 
             aes(x = age.base, y = height, color = "trees area 1"), 
             shape = 1) +
  # Points for Area 2
  geom_point(data = area2, 
             aes(x = age.base, y = height, color = "trees area 2"), 
             shape = 1) +
  # Lines for the models
  geom_line(data = res1.1, aes(x = x, y = y, color = "growth model a1")) +
  geom_line(data = res2.1, aes(x = x2, y = y, color = "growth model a2")) +
  # Manual colors
  scale_color_manual(
    name = "Legend",
    values = c(
      "trees area 1"      = "pink",
      "trees area 2"      = "grey",
      "growth model a1"   = "magenta",
      "growth model a2"   = "black"
    )
  ) +
  labs(
    x = "Age",
    y = "Height [m]"
  ) +
  theme_minimal()
```



